SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-12-07 back.tex

% [EB:] Removing this section title and moving relevant text to next section
%       We weren't talking about LA prefetchers here anyway, just SPP
% \subsection{Lookahead-based Prefetchers}
%% Simple stride prefetching techniques only detect sequences of addresses that differ by a constant value and fail to capture diverse memory access patterns. 
%Lookahead prefetchers can learn complicated patterns by collecting histories of observed data access patterns.
%EB: Didn't really understand this: 
%% "and correlating these with the strides, by considering each new page in a vacuum."

@@

Kim {\em et. al.} proposed Signature Path Prefetcher (SPP)~\cite{SPP}, 
a confidence-based lookahead prefetcher. 
SPP creates a signature associated with a page address by compressing the history of 
accesses. By correlating the signature with future likely delta patterns, SPP learns 
both simple and complicated memory access patterns quickly. 
While the basic idea of perceptron based prefetch filtering is applicable to
any look-ahead prefetcher, we develop a practical implementation of the idea
using SPP as our baseline. Here we describe the basic architecture of SPP.\\
%
\noindent \textbf{Signature Table:} As shown in figure \ref{fig:spp_structure}, 
the Signature Table keeps track of 256 most recently
accessed pages. It is meant to capture memory access patterns within a page
boundary. SPP indexes into an entry of the Signature Table using the page
number. For each entry corresponding to a page, the table stores a `last
block offset' and an `old signature'. Last block offset is the block offset
of the last memory access of that given page. The block offset is calculated
with respect to the page boundary. The signature is a 12-bit compressed
representation of the past few memory accesses for that page. The signature
is calculated as: $$New Signature = (\,Old Signature << 3 bits\,) \;\;XOR\;\;
(\,Delta\,)$$ Delta is the numerical difference between the block offset of
the current and the previous memory access. In case a matching page entry is
found, the stored signature retrieved and used to index into the Pattern
Table. This process is illustrated by figure \ref{fig:spp_update}.\\
%
\noindent \textbf{Pattern Table:} The Pattern Table is indexed by the signature
generated from the Signature Table. Pattern Table holds predicted delta
patterns and their confidence estimates. Each entry indexed by the signature
holds up to 4 unique delta predictions.\\
%

@@

%% NOTE: Can we do away with this example? 
%% If it looks like we are writing too much about SPP
%% EDIT: Decided to reorganize and remove this for now.
%% 	Consider the case that incoming page number 10 with a block offset 3
%% 	finds a match in the ST. The retrieved pattern signature is 0x30 and
%% 	the Last Offset is 1. Since now there is the Last Block Offset (1),
%% 	Incoming Block Offset (3), Old Signature (0x30), and New Signature
%% 	calculated as per above equation (0x182), SPP can infer
%% 	non-speculatively that the given pattern of memory accesses (as
%% 	captured in the Old Signature) leads to the particular Delta. In
%% 	general, Delta is defined as the difference between the prefetch
%% 	suggestion block and the initial block which triggered the prefetch.
%% 	In this case, since SPP is in learning phase, it is defined as the
%% 	difference between the Incoming Block Offset and the Last Block Offset
%% 	(+2 in this case). This, in turn generates the new memory pattern
%% 	(New Signature). This newly learned signature pattern and the delta
%% 	is stored in the Pattern Table.
\noindent \textbf{Prefetch Table:} The Prefetch Table is a 1024 entry 1-way associative
table that keeps a record of last few entries prefetched. This proves useful
for updating the prefetcher states if a tracked prefetch leads to a demand hit
or a cache eviction. \textit{Note:} The original paper refers to this as the
Prefetch Filter but since in our context filter refers to the perceptron
filter, we will be calling this structure as the Prefetch Table to avoid any
confusion.\\
%
\noindent \textbf{Lookahead Prefetching:} On each trigger, SPP goes down the program
speculation path using its own prefetch suggestion. Using current prefetch as
the baseline, it re-accesses the Pattern Table to generate further prefetches.
It repeats the cycle of accessing Pattern Table and updating the signature
based on highest confidence prefetch from the last iteration. The iteration
counter on which SPP manages to predict prefetch entries in the lookahead
manner is characterized as its `depth'. While doing so, SPP also keeps
compounding the confidence in each depth. Thus as depth increases, overall
confidence keeps decreasing.\\
%
\noindent \textbf{Confidence Tracking}: The Pattern Table keeps a count of hits to each
signature through a counter C\textsubscript{sig}. The number of hits for a
given delta per signature are tracked using a counter C\textsubscript{delta}.
The confidence for a given delta is approximated through C\textsubscript{d} =
C\textsubscript{delta} / C\textsubscript{sig}. When SPP enters into a
lookahead mode, the path confidence P\textsubscript{d} is given as:
$$P\textsubscript{d} \;=\; \alpha  \;.\;  C\textsubscript{d}  \;.\;
P\textsubscript{d-1}$$ Here $\alpha$ represents the global accuracy,
calculated as the ratio of the number of prefetches which led to a demand hit
to the number of prefetches recommended in total. The range of $\alpha$ is [0,1].
$D$ is the lookahead depth. For $d = 1$, when SPP is in
non-speculative mode, P\textsubscript{0} can be thought of as 1. 
The final P\textsubscript{d} is thresholded against prefetch threshold 
(T\textsubscript{p}) to reject the low confidence suggestions and 
then against a numerically bigger fill threshold (T\textsubscript{f}) to 
decide whether the prefetch should be sent to L2 Cache 
(high confidence prefetch) or Last Level Cache (low confidence prefetch).
The two thresholds were empirically set to 25 and 90 respectively, 
on the scale of 0 to 100.

>>>>

% [EB:] Removing this section title and moving relevant text to next section
%       We weren't talking about LA prefetchers here anyway, just SPP
% \subsection{Lookahead-based Prefetchers}
%% Simple stride prefetching techniques only detect sequences of addresses that differ by a constant value and fail to capture diverse memory access patterns. 
% Lookahead prefetchers can learn complicated patterns by collecting
% histories of observed data access patterns.  EB: Didn't really
% understand this:
%% "and correlating these with the strides, by considering each new page in a vacuum."

@@

Kim {\em et. al.} proposed Signature Path Prefetcher (SPP)~\cite{SPP},
a confidence-based lookahead prefetcher.  SPP creates a signature
associated with a page address by compressing the history of
accesses. By correlating the signature with future likely delta
patterns, SPP learns both simple and complicated memory access
patterns quickly.  While the basic idea of perceptron based prefetch
filtering is applicable to any look-ahead prefetcher, we develop a
practical implementation of the idea
using SPP as our baseline. Here we describe the basic architecture of SPP.

\noindent \textbf{Signature Table:} As shown in figure
\ref{fig:spp_structure}, the Signature Table keeps track of 256 most
recently accessed pages. It is meant to capture memory access patterns
within a page boundary. SPP indexes into an entry of the Signature
Table using the page number. For each entry corresponding to a page,
the table stores a `last block offset' and an `old signature'. Last
block offset is the block offset of the last memory access of that
given page. The block offset is calculated with respect to the page
boundary. The signature is a 12-bit compressed representation of the
past few memory accesses for that page. The signature is calculated
as:
$$New Signature = (\,Old Signature << 3 bits\,) \;\;XOR\;\;
(\,Delta\,)$$ Delta is the numerical difference between the block
offset of the current and the previous memory access. In case a
matching page entry is found, the stored signature retrieved and used
to index into the Pattern
Table. This process is illustrated by figure \ref{fig:spp_update}.

\noindent \textbf{Pattern Table:} The Pattern Table is indexed by the
signature generated from the Signature Table. Pattern Table holds
predicted delta patterns and their confidence estimates. Each entry
indexed by the signature holds up to 4 unique delta predictions.

@@

%% NOTE: Can we do away with this example? 
%% If it looks like we are writing too much about SPP
%% EDIT: Decided to reorganize and remove this for now.
%% 	Consider the case that incoming page number 10 with a block offset 3
%% 	finds a match in the ST. The retrieved pattern signature is 0x30 and
%% 	the Last Offset is 1. Since now there is the Last Block Offset (1),
%% 	Incoming Block Offset (3), Old Signature (0x30), and New Signature
%% 	calculated as per above equation (0x182), SPP can infer
%% 	non-speculatively that the given pattern of memory accesses (as
%% 	captured in the Old Signature) leads to the particular Delta. In
%% 	general, Delta is defined as the difference between the prefetch
%% 	suggestion block and the initial block which triggered the prefetch.
%% 	In this case, since SPP is in learning phase, it is defined as the
%% 	difference between the Incoming Block Offset and the Last Block Offset
%% 	(+2 in this case). This, in turn generates the new memory pattern
%% 	(New Signature). This newly learned signature pattern and the delta
%% 	is stored in the Pattern Table.
\noindent \textbf{Prefetch Table:} The Prefetch Table is a 1024 entry
1-way associative table that keeps a record of last few entries
prefetched. This proves useful for updating the prefetcher states if a
tracked prefetch leads to a demand hit or a cache
eviction. \textit{Note:} The original paper refers to this as the
Prefetch Filter but since in our context filter refers to the
perceptron filter, we will be calling this structure as the Prefetch
Table to avoid any confusion.

\noindent \textbf{Lookahead Prefetching:} On each trigger, SPP goes
down the program speculation path using its own prefetch
suggestion. Using current prefetch as the baseline, it re-accesses the
Pattern Table to generate further prefetches.  It repeats the cycle of
accessing Pattern Table and updating the signature based on highest
confidence prefetch from the last iteration. The iteration counter on
which SPP manages to predict prefetch entries in the lookahead manner
is characterized as its `depth'. While doing so, SPP also keeps
compounding the confidence in each depth. Thus as depth increases,
overall confidence keeps decreasing.

\noindent \textbf{Confidence Tracking}: The Pattern Table keeps a
count of hits to each signature through a counter
C\textsubscript{sig}. The number of hits for a given delta per
signature are tracked using a counter C\textsubscript{delta}.  The
confidence for a given delta is approximated through
C\textsubscript{d} = C\textsubscript{delta} /
C\textsubscript{sig}. When SPP enters into a lookahead mode, the path
confidence P\textsubscript{d} is given as:
$$P\textsubscript{d} \;=\; \alpha  \;.\;  C\textsubscript{d}  \;.\;
P\textsubscript{d-1}$$ Here $\alpha$ represents the global accuracy,
calculated as the ratio of the number of prefetches which led to a demand hit
to the number of prefetches recommended in total. The range of $\alpha$ is [0,1].
$D$ is the lookahead depth. For $d = 1$, when SPP is in
non-speculative mode, P\textsubscript{0} can be thought of as 1. 
The final P\textsubscript{d} is thresholded against prefetch threshold 
(T\textsubscript{p}) to reject the low confidence suggestions and 
then against a numerically bigger fill threshold (T\textsubscript{f}) to 
decide whether the prefetch should be sent to L2 Cache 
(high confidence prefetch) or Last Level Cache (low confidence prefetch).
The two thresholds were empirically set to 25 and 90 respectively, 
on the scale of 0 to 100.

### 2018-12-07 design_arch.tex

\begin{figure}
  \begin{center}
  \includegraphics[width=8cm]{PPF_Hierarchy}
    \label{fig:PPF_Hierarchy}
  \caption{PPF Architecture in the Memory Hierarchy}
  \end{center}
\end{figure}

@@

\begin{figure}[ht]
  \begin{center}
  \includegraphics[width=\columnwidth]{Datapath_Separate}
    \label{fig:PPF_Datapath}
  \caption{PPF Data Path and Operation}
  \end{center}
\end{figure}

@@

Figure~\ref{fig:PPF_Datapath} shows the required structures of PPF, as well as
the steps followed to filter out not-useful prefetches. The filter is
organized as a set of tables, where each entry in the tables holds a
\textit{weight}. For a configuration of PPF using $N$ number of features, $N$
different tables of weights are needed. Each feature is used to index a
distinct table. The number of entries of each table varies according to their
corresponding feature, hence, different number of bits are needed to index
different tables. Each weight is a 5-bit saturating counter ranging from -16
to +15. We found that having 5-bit weights provide a good trade-off between
accuracy and area. A detailed explanation of the storage overhead of PPF can
be found in Figure~\ref{tab:PPF_overhead}.\\

\noindent{\textbf{Inferencing}\\
%ETG:is this like an explanation of how prefetching happens? 
%EB: I see this as a short context of the datapath leading to the perc filter.
The base prefetcher is triggered on every demand access to the L2 Cache. At
this point, it has the opportunity to trigger a prefetch. If it decides to do
so, it will also need to decide how many cache blocks to prefetch. These
blocks can be either placed in the L2 or L3 cache according to the confidence
of the prefetching mechanism. Once the base prefetcher is triggered, the
suggested prefetch candidates are fed to the perceptron filter to determine
the usefulness of these prefetches. The filter makes the decision to issue
the prefetch suggestions from the base prefetcher. To make the decision, each
feature corresponding to a suggested prefetch is used to index a table and all
the corresponding weights are summed. The sum denotes the confidence value
for the suggested prefetch, and is thresholded against two different
magnitudes: PERC\_THRESHOLD\_HI and PERC\_ THRESHOLD\_ LO.

Prefetches whose sum exceeds PERC\_THRESHOLD\_HI are placed into the L2 cache.
The higher confidence value hints the prefetch would be useful and should be
prioritized. A prefetch for which the features result in a confidence value
between PERC\_THRESHOLD\_LO and PERC\_THRESHOLD\_ HI is allocated in the larger
LLC, as the filter is moderately confident of the future reuse of the cache
block, but not enough to possible pollute a significantly smaller L2.
Suggested prefetches for which the features lead to a confidence value lower
than PERC\_THRESHOLD\_LO are not prefetched, as the low confidence value
represents that the perceptron has learn that similar set of features are
associated with non-useful prefetches. \newline

@@

\noindent{\textbf{Training}\newline
The address from the demand request triggering the training is looked up in
both tables. If the address is in the prefetch table and marked as valid, this
hints the previous prediction was correct and this is a useful prefetch. We
compute the sum of the corresponding weights. If the sum falls below a
specific threshold, training occurs and the corresponding weights are adjusted
accordingly. These thresholds are introduced in order to avoid over-training,
helping the filter adapt quickly to changes in memory behavior. These
thresholds are referred to as POS\_UPDATE\_THRESHOLD ($\theta_p$) and \newline
NEG\_UPDATE\_THRESHOLD ($\theta_n$), respectively for the positive and
negative values of training saturation.

>>>>

\begin{figure}
  \begin{center}
  \includegraphics[width=8cm]{PPF_Hierarchy}
  \caption{PPF Architecture in the Memory Hierarchy}
  \label{fig:PPF_Hierarchy}
  \end{center}
\end{figure}

@@

\begin{figure}[ht]
  \begin{center}
  \includegraphics[width=\columnwidth]{Datapath_Separate}
  \caption{PPF Data Path and Operation}
  \label{fig:PPF_Datapath}
  \end{center}
\end{figure}

@@

Figure~\ref{fig:PPF_Datapath} shows the required structures of PPF, as well as
the steps followed to filter out not-useful prefetches. The filter is
organized as a set of tables, where each entry in the tables holds a
\textit{weight}. For a configuration of PPF using $N$ number of features, $N$
different tables of weights are needed. Each feature is used to index a
distinct table. The number of entries of each table varies according to their
corresponding feature, hence, different number of bits are needed to index
different tables. Each weight is a 5-bit saturating counter ranging from -16
to +15. We found that having 5-bit weights provide a good trade-off between
accuracy and area. A detailed explanation of the storage overhead of PPF can
be found in Section~\ref{Method-Overheads}.\\

\noindent{\textbf{Inferencing}\\
%ETG:is this like an explanation of how prefetching happens? 
%EB: I see this as a short context of the datapath leading to the perc filter.
The base prefetcher is triggered on every demand access to the L2 Cache. At
this point, it has the opportunity to trigger a prefetch. If it decides to do
so, it will also need to decide how many cache blocks to prefetch. These
blocks can be either placed in the L2 or L3 cache according to the confidence
of the prefetching mechanism. Once the base prefetcher is triggered, the
suggested prefetch candidates are fed to the perceptron filter to determine
the usefulness of these prefetches. The filter makes the decision to issue
the prefetch suggestions from the base prefetcher. To make the decision, each
feature corresponding to a suggested prefetch is used to index a table and all
the corresponding weights are summed. The sum denotes the confidence value
for the suggested prefetch, and is thresholded against two different
magnitudes: $\tau_{hi}$ and $\tau_{lo}$.

Prefetches whose sum exceeds $\tau_{hi}$ are placed into the L2 cache.
The higher confidence value hints the prefetch would be useful and should be
prioritized. A prefetch for which the features result in a confidence value
between $\tau_{lo}$ and $\tau_{hi}$ is allocated in the larger
LLC, as the filter is moderately confident of the future reuse of the cache
block, but not enough to possible pollute a significantly smaller L2.
Suggested prefetches for which the features lead to a confidence value lower
than $\tau_{lo}$ are not prefetched, as the low confidence value
represents that the perceptron has learn that similar set of features are
associated with non-useful prefetches. \newline

@@

\noindent{\textbf{Training}\newline
The address from the demand request triggering the training is looked up in
both tables. If the address is in the prefetch table and marked as valid, this
hints the previous prediction was correct and this is a useful prefetch. We
compute the sum of the corresponding weights. If the sum falls below a
specific threshold, training occurs and the corresponding weights are adjusted
accordingly. These thresholds are introduced in order to avoid over-training,
helping the filter adapt quickly to changes in memory behavior. These
thresholds are referred to as $\theta_p$ and $\theta_n$, 
respectively for the positive and negative values of training saturation.

### 2018-12-07 intro.tex

\begin{figure}[t]
\includegraphics[width=\columnwidth]{Motivation}
\caption{Variation of IPC, Total Prefetches and Useful Prefetches with
  lookahead depth.}
\label{Fig:Motivation}
\end{figure}

@@

In a single core configuration, PPF increases performance by 6.84\%
compared to the underlying prefetcher, SPP. In a multi-core system
running a mixes of memory intensive SPEC CPU 2017 traces, PPF saw an
improvement of 11.9\% over SPP for a 4-core system, and 10.67\% for an
8-core system.  \pg{update numbers before submit}

>>>>

\begin{figure}[t]
\includegraphics[width=\columnwidth]{Motivation}
\caption{Impact of aggressive prefetching on performance. X-asis
  represents increasingly aggressive prefetch lookahead depth. All the
  metrics have been normalized to the least aggressive
  configuration. As can be seen, total number of prefetches increases
  at a higher rate than the number of good prefetches, leading to IPC
  reduction.}
\label{Fig:Motivation}
\end{figure}

@@

In a single core configuration, PPF increases performance by 3.78\%
compared to the underlying prefetcher, SPP. In a multi-core system running a
mixes of memory intensive SPEC CPU 2017 traces, PPF saw an improvement
of 11.4\% over SPP for a 4-core system, and 9.65\% for an 8-core system.

### 2018-12-07 method.tex

%This is the first time that SPEC CPU 2017 benchmark suite~\cite{SPEC2017}
%has been used to characterize and measure the prefetch performance.
We use the SPEC CPU 2017 benchmark suite~\cite{SPEC2017}. We use all the 20
workloads available in the SPEC CPU 2017 suite. Using the SimPoint~\cite{SimPoint}
methodology, we identified 95 different program segments of 1 Billion
instructions each.

\noindent \textit{Single-core performance:} For single-core simulations, we use 200
million instructions to warm-up the microarchitectural structures and the next
one billion instructions to do detailed simulations and collect run-time
statistics. We report the IPC speedup over the baseline of no prefetching.
The final numbers reported are the geometric mean of the weighted mean speedup 
achieved per application using the SimPoint methodology.\\

\noindent \textit{Multi-core performance:} For multi-application workloads, we generate
100 random mixes and another 100 mixes from the memory intensive subset of
SPEC CPU 2017. For 4-core workload, 200 Million instructions are used for warm-up
and additional 1 Billion instruction simulated for collecting statistics.
Each CPU keeps executing its workload till the last CPU completes one billion
instructions after warm-up. For collecting IPC and other data, only the first
billion instructions are considered as the region of interest.

@@

We repeat the same process for 8-core workloads, correspondingly with 16MB
LLC. The only difference is that 20 million warm-up instructions and 100
million full instructions are executed. This is done so as to keep the
simulation run-time within reasonable limits as a single 8-core mix takes up
to 3 days to simulate one billion instructions.\\

\noindent \textit{Validation:} We cross-validated our PPF model using SPEC
CPU 2006~\cite{SPEC2006} and CloudSuite~\cite{CloudSuite} benchmarks. For
single-core SPEC CPU 2006, we developed 94 simpoints spread across all the 29
applications. For multi-core, we followed the same methodology as SPEC CPU 2017.
For CloudSuite, we used the traces made available for the 2nd Cache
Replacement Competition (CRC-2)~\cite{CRC_2}. The traces include 4 4-core
applications with 6 distinct phases per application.

@@

In this section, we analyze the hardware overhead required to
implement PPF. The Prefetch Table was enhanced to accommodate
storing of meta-data for perceptron training. Table
\ref{tab:PTable_metadata} depicts the meta-data stored for each entry in
the Prefetch Table. Table \ref{tab:PPF_overhead} shows the total
storage overhead of PPF implementation. The hardware budget for
2nd Data Prefetching championship was 32 KB. Keeping that in mind 
the considerable speedup PPF obtained over the winner, the extra hardware
budget can be accounted for. The extra hardware also makes the
overall scheme more scalable than SPP. In the original SPP paper, it
was demonstrated that adding extra hardware brings little advantage in
terms of performance gain. The newly added perceptron tables can be
scaled to increase / decrease features depending on the permitted
budget.

>>>>

%This is the first time that SPEC CPU 2017 benchmark suite~\cite{SPEC2017}
%has been used to characterize and measure the prefetch performance.
We use the SPEC CPU 2017 benchmark suite~\cite{SPEC2017}. We use all the 20
workloads available in the SPEC CPU 2017 suite. Using the SimPoint~\cite{SimPoint}
methodology, we identified 95 different program segments of 1 Billion
instructions each.\\
%
\noindent \textit{Single-core performance:} For single-core simulations, we use 200
million instructions to warm-up the microarchitectural structures and the next
one billion instructions to do detailed simulations and collect run-time
statistics. We report the IPC speedup over the baseline of no prefetching.
The final numbers reported are the geometric mean of the weighted mean speedup 
achieved per application using the SimPoint methodology.\\
%
\noindent \textit{Multi-core performance:} For multi-application workloads, we generate
100 random mixes and another 100 mixes from the memory intensive subset of
SPEC CPU 2017. For 4-core workload, 200 Million instructions are used for warm-up
and additional 1 Billion instruction simulated for collecting statistics.
Each CPU keeps executing its workload till the last CPU completes one billion
instructions after warm-up. For collecting IPC and other data, only the first
billion instructions are considered as the region of interest.

@@

We repeat the same process for 8-core workloads, correspondingly with 16MB
LLC. The only difference is that 20 million warm-up instructions and 100
million full instructions are executed. This is done so as to keep the
simulation run-time within reasonable limits as a single 8-core mix takes up
to 3 days to simulate one billion instructions.\\
%
\noindent \textit{Validation:} We cross-validated our PPF model using SPEC
CPU 2006~\cite{SPEC2006} and CloudSuite~\cite{CloudSuite} benchmarks. For
single-core SPEC CPU 2006, we developed 94 simpoints spread across all the 29
applications. For multi-core, we followed the same methodology as SPEC CPU 2017.
For CloudSuite, we used the traces made available for the 2nd Cache
Replacement Competition (CRC-2)~\cite{CRC_2}. The traces include 4 4-core
applications with 6 distinct phases per application.

@@

\subsection{Overhead for PPF}
\label{Method-Overheads}
In this section, we analyze the hardware overhead required to
implement PPF. The Prefetch Table was enhanced to accommodate
storing of meta-data for perceptron training. Table
\ref{tab:PTable_metadata} depicts the meta-data stored for each entry in
the Prefetch Table. Table \ref{tab:PPF_overhead} shows the total
storage overhead of PPF implementation. The hardware budget for
2nd Data Prefetching championship was 32 KB. Keeping that in mind 
the considerable speedup PPF obtained over the winner, the extra hardware
budget can be accounted for. The extra hardware also makes the
overall scheme more scalable than SPP. In the original SPP paper, it
was demonstrated that adding extra hardware brings little advantage in
terms of performance gain. The newly added perceptron tables can be
scaled to increase / decrease features depending on the permitted
budget.
